Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-05-03 overleaf/vectorization.tex
Done with backend.

{\small
\begin{pythonn}
for i in range(0, num_bins):
  res1 = PHI(res, res2)
  for j in range(0, N):
    res2 = PHI(res1, res3)
    tmp1 = (A[j] == i)
    tmp2 = (res2[i] + B[j])
    tmp3 = MUX(tmp1, res2[i], tmp2)
    res3 = Update(res2, i, tmp3)
return res1
\end{pythonn}
}

There is loop-carried dependence from 8 to 2 for write array {\sf res} even though the loop can be vectorized across $i$.

@@

The following algorithm removes certain infeasible loop-carried dependencies that are due to array writes. Consider a loop with index $0 \le j < J$
nested at $i,j,k$. 
%Here $i$ represents the enclosing loops of $j$ and $k$ represents the enclosed loops in $j$.
\begin{comment}
\begin{algorithmic}
\FOR {each array ${\sf A}$ written in loop $j$}
\STATE \COMMENT { including enclosed loops in $j$ }
\STATE dep = False
\FOR {each pair def: ${\sf A}_m[f(i,j,k)] = ... $, and use: $ ... = {\sf A}_n[f'(i,j,k)]$ in loop $j$}
\IF {$\exists \underline{i}, \underline{j}, \underline{j}', \underline{k}, \underline{k}'$, s.t. $0\! \le\! \underline{i}\! <\! I,\: 0\! \le\! \underline{j}, \underline{j}' \! <\! J,\: 0\! \le\! \underline{k}, \underline{k}' \! <\! K,\: \underline{j}<\underline{j}'$, and $f(\underline{i},\underline{j},\underline{k}) = f'(\underline{i},\underline{j}',\underline{k}')$}
\STATE dep = True
\ENDIF
\ENDFOR
\IF {dep == False}
\STATE remove back edge into ${\sf A}$'s $\phi$-node in loop $j$.
\ENDIF
\ENDFOR
\end{algorithmic}
\end{comment}
Consider a loop $j$ enclosed in some fixed $\underline{i}$. Only if an update (definition) ${\sf A}_m[f(i,j,k)] = ... $ at some iteration $\underline{j}$
references the \emph{same} array element as a use $ ... = {\sf A}_n[f'(i,j,k)]$ at some later iteration $\underline{j}'$,
we may have a loop-carried dependence for ${\sf A}$ due to this def-use pair. (In contrast, Cytron's algorithm inserts a loop-carried dependency every time there is an array update.)
The algorithm, shown in~\cite{Anon_TR}, examines all def-use pairs in loop $j$, including defs and uses in nested loops, 
searching for values $\underline{i}, \underline{j}, \underline{j}', \underline{k}, \underline{k}'$ that satisfy
$f(\underline{i},\underline{j},\underline{k}) = f'(\underline{i},\underline{j}',\underline{k}')$ (using Z3). If such values exist for some def-use pair, then there is a potential
loop-carried dependence on ${\sf A}$; otherwise there is not and we can remove the spurious backward edge to the array PHI-node 
thus ``freeing up'' statements for vectorization. %\ana{Double check this!!!}

In Histogram, this removes the back edge from the update at 8 to the PHI node at 2. Clearly there does not exist $i < i'$ that makes
$i$ at 9 and $i'$ at 6 and 7 equal. The back-edge from 9 to 4 stays because for every $j < j'$ 9 and 6 and 7 access the same location in {\sf res}. %\ana{Can't wrap my head around this...}

@@

Note however, that this step renders some array phi-nodes target-less. We handle target-less phi-nodes with a minor extension of Vectorization (Phase 2).
First, we merge closures that update the same array. This simplifies handling of array $\phi$-nodes: if each closure is turned into a separate loop
each loop will need to have its own array phi-node to account for the update and this would complicate the analysis.
Second, we add the target-less node of array {\sf A} back to the closure that updates {\sf A} ---
the intuition is, even if there is no loop-carried dependence from writes to reads on {\sf A}, {\sf A} is written and the write (i.e., update) cannot be vectorized;
therefore, the updated array has to carry to the next iteration of the loop. Third, in cases when the phi-node remains target-less, i.e., cases when the
array write can be vectorized, we have to properly remove the phi-node replacing uses of the left-hand side of the phi-node with its arguments.
Additional detail, as well as detailed treatment of the minor changes to $\mathit{raise\_dim}$ and $\mathit{drop\_dim}$ required to do expansion of arrays 
in addition to scalars, appear in~\cite{Anon_TR}.

@@

res1^ = raise_dim(res, (j:N)) // replacing res1 with res, 1st arg
for j in range(0, N):
   res2 = PHI(res1^, res3)
   tmp2[I,j] = (res2[I,j] + B1[I,j])
   tmp3[I,j] = MUX(tmp1[I,j], res2[I,j], tmp2[I,j])
   res3 = Update(res2, (I,j), tmp3)
   equiv. to res3 = res2; res3[I,j] = tmp3[I,j]
res2^ = drop_dim(res2)
res1 = drop_dim(res2^) // replacing with res2^, 2nd arg. NOOP
return res1
\end{pythonn}
}

>>>>

{\small
\begin{pythonn}
for i in range(0, num_bins):
  res1 = PHI(res, res2) # res is the result array
  for j in range(0, N):
    res2 = PHI(res1, res3)
    tmp1 = (A[j] == i)
    tmp2 = (res2[i] + B[j])
    tmp3 = MUX(tmp1, res2[i], tmp2)
    res3 = Update(res2, i, tmp3)
return res1
\end{pythonn}
}

There is a def-use edge to 2 due to the update at 8 of array {\sf res} even though the loop can be vectorized across $i$.

@@

The following algorithm removes certain infeasible loop-carried dependencies that are due to array writes. Consider a loop with index $0 \le j < J$
nested at $i,j,k$. 
%Here $i$ represents the enclosing loops of $j$ and $k$ represents the enclosed loops in $j$.
\begin{comment}
\begin{algorithmic}
\FOR {each array ${\sf A}$ written in loop $j$}
\STATE \COMMENT { including enclosed loops in $j$ }
\STATE dep = False
\FOR {each pair def: ${\sf A}_m[f(i,j,k)] = ... $, and use: $ ... = {\sf A}_n[f'(i,j,k)]$ in loop $j$}
\IF {$\exists \underline{i}, \underline{j}, \underline{j}', \underline{k}, \underline{k}'$, s.t. $0\! \le\! \underline{i}\! <\! I,\: 0\! \le\! \underline{j}, \underline{j}' \! <\! J,\: 0\! \le\! \underline{k}, \underline{k}' \! <\! K,\: \underline{j}<\underline{j}'$, and $f(\underline{i},\underline{j},\underline{k}) = f'(\underline{i},\underline{j}',\underline{k}')$}
\STATE dep = True
\ENDIF
\ENDFOR
\IF {dep == False}
\STATE remove back edge into ${\sf A}$'s $\phi$-node in loop $j$.
\ENDIF
\ENDFOR
\end{algorithmic}
\end{comment}
Consider a loop $j$ enclosed in some fixed $\underline{i}$. Only if an update (definition) ${\sf A}_m[f(i,j,k)] = ... $ at some iteration $\underline{j}$
references the \emph{same} array element as a use $ ... = {\sf A}_n[f'(i,j,k)]$ at some later iteration $\underline{j}'$,
we may have a loop-carried dependence for ${\sf A}$ due to this def-use pair. (In contrast, Cytron's algorithm inserts a loop-carried dependency every time there is an array update.)
The algorithm, shown in~\cite{Anon_TR}, examines all def-use pairs in loop $j$, including defs and uses in nested loops, 
searching for values $\underline{i}, \underline{j}, \underline{j}', \underline{k}, \underline{k}'$ that satisfy
$f(\underline{i},\underline{j},\underline{k}) = f'(\underline{i},\underline{j}',\underline{k}')$ (using Z3). If such values exist for some def-use pair, then there is a potential
loop-carried dependence on ${\sf A}$; otherwise there is not and we can remove the spurious backward edge to the array PHI-node 
for loop $j$, thus ``freeing up'' statements for vectorization. %\ana{Double check this!!!}

In Histogram, this removes the back edge from 4 to the PHI node at 2. Clearly there does not exist $i < i'$ that makes
$i$ at 9 and $i'$ at 6 and 7 equal. The back-edge from 9 to 4 stays because for every $j < j'$ 9 and 6 and 7 access the same location in {\sf res}. %\ana{Can't wrap my head around this...}

@@

Removing infeasible edges renders some array phi-nodes \emph{target-less}. We handle target-less phi-nodes with a minor extension of Vectorization (Phase 2).
First, we merge closures that update the same array. This simplifies handling of array $\phi$-nodes: if each closure is turned into a separate loop
each loop will need to have its own array phi-node to account for the update and this would complicate the analysis.
Second, we add the target-less node of array {\sf A} back to the closure that updates {\sf A} ---
the intuition is, even if there is no loop-carried dependence from writes to reads on {\sf A}, {\sf A} is written and the write (i.e., update) cannot be vectorized;
therefore, the updated array has to carry to the next iteration of the loop. Third, in cases when the phi-node remains target-less, i.e., cases when the
array write can be vectorized, we have to properly remove the phi-node replacing uses of the left-hand side of the phi-node with its arguments.
Additional detail on the minor changes to $\mathit{raise\_dim}$ and $\mathit{drop\_dim}$ required to do expansion of arrays appear in~\cite{Anon_TR}.

@@

res1^ = raise_dim(res, (j:N)) # replacing res1 with res, 1st arg
for j in range(0, N):
   res2 = PHI(res1^, res3)
   tmp2[I,j] = (res2[I,j] + B1[I,j])
   tmp3[I,j] = MUX(tmp1[I,j], res2[I,j], tmp2[I,j])
   res3 = Update(res2, (I,j), tmp3)
   equiv. to res3 = res2; res3[I,j] = tmp3[I,j]
res2^ = drop_dim(res2)
res1 = drop_dim(res2^) # replacing with res2^, 2nd arg. NOOP
return res1
\end{pythonn}
}
